Assessing landscape N removal in coastal New England catchments using the

### 2023-06-07 manuscript/manuscript.Rmd
Add packages for figs 4 and 5. Add new code for fig 5 and table 3, which accompanies fig 5.

---
title: Assessing landscape N removal in coastal New England catchments using the N-Sink
  approach with a new R package, nsink
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water quality. Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery. Natural features within watersheds can act as landscape sinks for N, such as wetlands, streams and ponds that transform dissolved N into gaseous N, effectively removing it from the aquatic system. Identifying and evaluating these landscape sinks and their spatial relationship to N sources can help managers understand the effects of alternative decisions on downstream resources. N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 (or larger) catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment. A series of static maps are produced to visualize N removal efficiency, transport and delivery, the latter in the form of an index. To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and applied to produce static maps for all HUC-12 catchments in southern New England. The R package can be used to develop N delivery maps for a specific HUC-12 of interest, or users can visit the University of Connecticut website to explore the already mapped areas. Users can also investigate specific flowpaths interactively by clicking on any location within the catchment. A flowpath is generated, along with a table describing N removal along each segment.
---

@@

```{r setup, echo=FALSE, warning = FALSE}
library(FedData)
library(nsink)
library(raster)
```

@@

```{r prep-data, eval = FALSE, include = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072

@@

Users can use the following to interactively identify a starting point for a flowpath of interest. 
```{r case1, echo = TRUE, eval = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072
# Select location on map for starting point using cursor and click
pt <- unlist(locator(n = 1))
# Convert to sf POINT
start_loc <- st_sf(st_sfc(st_point(pt), crs = aea))
# Generate the flowpath
niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
# Summarize N removal along the flowpath
niantic_removal <- nsink_calc_removal(niantic_data)
# table of removal within flowpath segments
niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) 
```

@@

ptb <- c(1953529, 2290811)
start_loc_inter_b <- st_sf(st_sfc(st_point(ptb), crs = aea))

niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)
raster::plot(niantic_static_maps$transport_idx, breaks = c(0, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100), 
             col = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", "#E2F088", "#F5E871",
                     "#F9C159", "#F99844", "#F56A2E","#EF2820"))
huc <- niantic_data$huc
streams <- st_intersection(huc, niantic_data$streams)
plot(streams, add = TRUE, col = "dodgerblue")
plot(niantic_data$lakes, add = TRUE, col = "dodgerblue", border = "dodgerblue")
plot(st_geometry(niantic_fp_a$flowpath_network), add = TRUE, col = "gray40", lwd = 3.0)
plot(st_geometry(niantic_fp_a$flowpath_ends), add = TRUE, col = "gray40", lwd = 3.0)
plot(st_geometry(niantic_fp_b$flowpath_network), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
plot(st_geometry(niantic_fp_b$flowpath_ends), add = TRUE, col = "gray0", lwd = 2.0, lty = 2)
x <- c(pta[1], ptb[1])
y <- c(pta[2], ptb[2])
ab <- st_multipoint(matrix(c(x,y), nrow = 2, ncol = 2))
ab_plot <- st_sf(st_sfc(ab), crs = aea)
plot(ab_plot, add = TRUE, pch = 16)
text(ab, labels = c("A", "B"), pos = 1, offset = 0.4)
```

@@

Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps. The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment. Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled. Figure 5B shows the N Delivery Index from these two developed areas. Note that the northern area is less densely developed than the southern area but the N Delivery Index is higher in the northern area as compared to the southern area, suggesting that N source controls within the northern development might be considered.

```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two developed areas."}
box <- c(1949902, 1953930, 2288202, 2292262)
e <- extent(box)
legend_nlcd <- pal_nlcd()
case2 <- crop(niantic_data$nlcd, e)
vals <- unique(case2@data@values)
leg <- legend_nlcd[legend_nlcd$ID %in% vals,]
case2_north <- st_sf(st_sfc(st_point(c(1950846, 2290912)), crs = aea)) 
case2_south <- st_sf(st_sfc(st_point(c(1952693, 2289249)), crs = aea))

my_key <- list(rectangles=list(col = leg$Color),
              text=list(lab=leg$Class),
              space='left',
              columns=1,
              size=2,
              cex=.6)
colors <- legend_nlcd$Color
plot(case2, col = legend_nlcd$Color)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)
```
```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index over land cover, with hotter colors denoting higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency."}
case2_delivery <- crop(niantic_static_maps$delivery_idx, e)
plot(case2, col = legend_nlcd$Color, alpha = 0.6, legend = FALSE)
plot(case2_delivery, breaks = c(20, 40, 60, 80, 100), col = c("#FFBEBE", "#F57A7A","#A80405","#652600"), add = TRUE)
plot(case2_north, pch = 1, col = "black", cex = 25, lwd = 3, add = TRUE)
plot(case2_south, pch = 1, col = "black", cex = 18, lwd = 3, add = TRUE)

```

>>>>

---
title: "Assessing landscape N removal in coastal New England catchments using the
  N-Sink approach with a new R package, nsink"
author:
- Kellogg, D. Q. ^*1*^, Hollister. J. W. ^\*^ ^*2*^, Arnold, C.^*3*^, Gold, A.^*1*^,
  Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K. J.^*4*^
- ^*1*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*2*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*3*^ University of Connecticut, Center for Land Use Education and Research, Haddam,
  CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development, Groundwater
  Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: "Excess nitrogen (N) loading to coastal ecosystems impairs estuarine water
  quality. Land management decisions made within estuarine watersheds have a direct
  impact on downstream N delivery. Natural features within watersheds can act as landscape
  sinks for N, such as wetlands, streams and ponds that transform dissolved N into
  gaseous N, effectively removing it from the aquatic system. Identifying and evaluating
  these landscape sinks and their spatial relationship to N sources can help managers
  understand the effects of alternative decisions on downstream resources. N-Sink
  is an approach that uses widely available GIS data to identify landscape sinks within
  HUC-12 (or larger) catchments, estimate their N removal potential and map the effect
  of those sinks on N movement through the catchment. A series of static maps are
  produced to visualize N removal efficiency, transport and delivery, the latter in
  the form of an index. To facilitate data acquisition, processing and visualization,
  an R package `nsink` was developed and applied to produce static maps for all HUC-12
  catchments in southern New England. The R package can be used to develop N delivery
  maps for a specific HUC-12 of interest, or users can visit the University of Connecticut
  website to explore the already mapped areas. Users can also investigate specific
  flowpaths interactively by clicking on any location within the catchment. A flowpath
  is generated, along with a table describing N removal along each segment."
---

@@

```{r setup, echo=FALSE, warning = FALSE}
library(FedData)
library(nsink)
library(raster)
library(ggplot2)
library(terra)
library(tidyterra)
```

@@

```{r prep-data, eval = TRUE, include = FALSE}
# EPSG for CONUS Albers Equal Area
aea <- 5072

@@

Users can use the following to interactively identify a starting point for a flowpath of interest. 
```{r case1, echo = TRUE, eval = FALSE}

# Select location on map for starting point using cursor and click
pt <- unlist(locator(n = 1))
# Convert to sf POINT
start_loc <- st_sf(st_sfc(st_point(pt), crs = aea))
# Generate the flowpath
niantic_fp <- nsink_generate_flowpath(start_loc, niantic_data)
# Summarize N removal along the flowpath
niantic_removal <- nsink_calc_removal(niantic_data)
# table of removal within flowpath segments
niantic_fp_removal <- nsink_summarize_flowpath(niantic_fp, niantic_removal) 
```

@@

ptb <- c(1953529, 2290811)
start_loc_inter_b <- st_sf(st_sfc(st_point(ptb), crs = aea))
niantic_fp_b <- nsink_generate_flowpath(start_loc_inter_b, niantic_data)
niantic_fp_removal_b <- nsink_summarize_flowpath(niantic_fp_b, niantic_removal)

huc <- niantic_data$huc
streams <- st_intersection(huc, niantic_data$streams)
x <- c(pta[1], ptb[1])
y <- c(pta[2], ptb[2])
labels <- c("A", "B")
pts_df <- data.frame(x,y,labels)
ab_plot <- st_as_sf(pts_df, coords = c("x","y"), crs = aea)

from <- c(seq(0,90,10), NA)
to <- c(seq(10,100,10), NA)

tnsport_idx <- classify(rast(niantic_static_maps$transport_idx),matrix(c(from,to,to),ncol = 3))
values(tnsport_idx) <- factor(values(tnsport_idx))

ggplot() +
  geom_spatraster(data = tnsport_idx) +
  scale_fill_manual(values = c("#38A1D0", "#7AB4C0", "#A2C8B0","#C3DC9D", 
                                        "#E2F088", "#F5E871", "#F9C159", 
                                        "#F99844", "#F56A2E","#EF2820"), 
                                        na.value = NA, na.translate = FALSE) +
  geom_sf(data = streams, color = "dodgerblue", linewidth = 0.7) +
  geom_sf(data = niantic_data$lakes, fill = "dodgerblue", color = "dodgerblue") +
  geom_sf(data = niantic_fp_a$flowpath_network, color = "grey40", linewidth = 2.5) +
  geom_sf(data = niantic_fp_a$flowpath_ends, color = "grey40", linewidth = 2.5) +
  geom_sf(data = niantic_fp_b$flowpath_network, color = "grey0", linewidth = 0.9) +
  geom_sf(data = niantic_fp_b$flowpath_ends, color = "grey0", linewidth = 0.9) +
  geom_sf(data = ab_plot, pch = 16, size = 2) + 
  geom_sf_label(data = ab_plot, aes(label = labels))
```

@@

Similarly, sources of N that are contributing relatively more N due to their location within the catchment can be highlighted using the `nsink` static maps. The N Delivery Index map combines N sources with N transport to highlight developed areas that are located on "leakier parts" of a catchment. Figure 5A displays a high resolution image of land cover in the same watershed as Example 1, with two developed areas circled.

```{r plot5a, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5A: Land cover within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas (Table 3)."}

box <- c(1950000, 1953800, 2289700, 2293100)
e <- ext(box)
legend_nlcd <- pal_nlcd()
case2 <- crop(rast(niantic_data$nlcd), e)
vals <- unique(case2$Class)
leg <- legend_nlcd[legend_nlcd$ID %in% vals$Class,]
values(case2) <- factor(values(case2))

bdist <- units::set_units(0.5, km)
ptw <- st_point(x = c(1951000, 2290650))
ptw <- ptw %>% st_sfc(crs = aea)
ptw_buf <- st_buffer(ptw, bdist)

pte <- st_point(x = c(1952900, 2291750))
pte <- pte %>% st_sfc(crs = aea)
pte_buf <- st_buffer(pte, bdist)

ggplot() +  geom_spatraster(data = case2, aes(fill = Class)) +
  scale_fill_manual(values = leg$Color, na.value = NA, na.translate = FALSE,
                    labels = leg$Class) + labs(fill = "NLCD Class") +
  theme(axis.title.x = element_blank(), axis.title.y = element_blank(),
        legend.text = element_text(size = 7), 
        legend.title = element_text(size = 10)) +
  geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
```

```{r table3-build, eval = TRUE, include = FALSE}
table3 <- data.frame("Land Cover" = c("Developed", "Agriculture", "Forest/Vegetated",
                                      "Wetlands", "Other"),
                     "West (%)" = c("23","7","64","6","0"),
                     "East (%)" = c("20","0","61","18","1"),
                     check.names = FALSE)
```
```{r table3, include = TRUE, echo = FALSE}
knitr::kable(table3, caption = "Table 3: Summary of Land Cover in Highlighted Areas of Interest")
```

An analysis of N delivery from these two similarly developed areas, however, show the western area is leakier than the eastern area, as assessed by summing the delivery index values (Figure 5B; West = 15759, East = 6770), suggesting that N source controls within the western development might be considered.

```{r, plot5b, include = TRUE, echo = FALSE, warning = FALSE, fig.cap = "Figure 5B: Nitrogen Delivery Index within an area of interest in the Niantic River HUC-12, highlighting two similarly developed areas. Hotter colors denote higher N delivery based on normalized loading from different land covers multiplied by the N Transport Efficiency."}

from <- c(seq(0,90,10), NA)
to <- c(seq(10,100,10), NA)
delivery_idx <- classify(rast(niantic_static_maps$delivery_idx),matrix(c(from,to,to),ncol = 3))
case2_delivery <- crop(delivery_idx, e)
case2_delivery[case2_delivery < 20] <- 10
values(case2_delivery) <- factor(values(case2_delivery))

cols10 <-  c("white", "#FFBEBE", "#FAA8A8", "#FA9191", "#F57A7A",
           "#DB5252", "#C22B2C", "#A80405",
           "#910F04", "#7B1B02", "#652600")

ggplot() + geom_spatraster(data = case2_delivery) +
  scale_fill_manual(values = cols10, na.value = "white", na.translate = FALSE) +
  guides(fill = guide_legend(reverse = TRUE)) + 
  labs(fill = "") +
  geom_sf(data = ptw_buf, color = "black", fill = NA, linewidth = 1.5) +
  geom_sf(data = pte_buf, color = "black", fill = NA, linewidth = 1.5)
```
